Theory of surface second-harmonic generation for semiconductors including effects of nonlocal operators

### 2014-11-21 paper-v-0.tex
finished correcting equations

\subsection{Length Gauge Formalism}
To calculate the optical response 
within the Length Gauge formalism, we take the 
interaction Hamiltonian as
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).\label{rde}
\end{equation}
The 
treatment of the position operator
$\mathbf{r}$ for extended Bloch states, $\langle\mathbf{r}|n\mathbf{k}\rangle$ has 
been described in detail.\cite{adamsJCP53,blountSSP62} 
In particular, Ref.~\onlinecite{aversaPRB95} develops
the correct treatment of $\mathbf{r}$ for
the calculation of the nonlinear optical response.
$\mathbf{r}$ is split into the {\it intraband} 
($\mathbf{r}_i$) and {\it interband} ($\mathbf{r}_e$) parts, where 
$\mathbf{r}=\mathbf{r}_i+\mathbf{r}_e$. 
For $\mathbf{r}_e$ one uses
\begin{align}
\langle n\mathbf{k} | \mathbf{r}_e | m\mathbf{k}'\rangle &=
(1-\delta_{nm})\delta(\mathbf{k}-\mathbf{k}')\boldsymbol{\xi}_{nm}(\mathbf{k})
,
\label{rnminn}
\end{align}
such that $\mathbf{r}_{e,nm}=0$ for $n=m$.
From Eq.~\eqref{mv} with $H_0\to
H^\Sigma_0$, we obtain
\begin{align}
\mathbf{r}_{e,nm}(\mathbf{k}) =
\boldsymbol{\xi}_{nm}(\mathbf{k})\equiv 
\mathbf{r}_{nm}(\mathbf{k}) 
&=
\frac{\mathbf{v}^\Sigma_{nm}(\mathbf{k})}{i\omega^\Sigma_{nm}(\mathbf{k})}
\quad\quad n\notin D_m 
,
\label{pmnrmn}
\end{align}  
where we defined
$\omega^\Sigma_{nm}(\mathbf{k})\equiv\omega^\Sigma_n(\mathbf{k})-\omega^\Sigma_m(\mathbf{k})$, and
$D_m$ are all the possible degenerate $m$-states. 
%We recognize that for each instance in which 
%$n\notin D_m$ or $n\ne m$, $\mathbf{r}_{nm}(\mathbf{k})$ are interband matrix
%elements.
For the intraband part, $\mathbf{r}_i$ only appears in
commutators during the derivation of
the optical response. We use,\cite{aversaPRB95}
\begin{equation}
\langle n\mathbf{k} | [\mathbf{r}_i,{\cal O}] | m\mathbf{k}'\rangle
=i\delta(\mathbf{k}-\mathbf{k}')({\cal O}_{nm})_{;\mathbf{k}},
\label{conmri3n}
\end{equation}  
where
\begin{equation}
({\cal O}_{nm})_{;\mathbf{k}}=
\nabla_\mathbf{k} 
{\cal O}_{n m}(\mathbf{k}) 
- 
i 
{\cal O}_{nm}(\mathbf{k}) 
\left(
\boldsymbol{\xi}_{nn}(\mathbf{k}) 
-
\boldsymbol{\xi}_{mm}(\mathbf{k}) 
\right) 
,
\label{gendevnn}
\end{equation} 
is
the generalized derivative 
of the operator ${\cal O}$. 
The vectors $\boldsymbol{\xi}_{nn}(\mathbf{k})$ are defined in 
Ref.~\onlinecite{aversaPRB95} though they do not need to be 
calculated explicitly in what follows.

@@

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate along the crystallographic [011] direction is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{images/si2x1-crop}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{equation}
\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega) 
,
\end{equation}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ through the full slab.
We show the results for this comparison in the remainder 
of this section.

@@

From Eqs.~\eqref{lab} and \eqref{vnl} 
\begin{equation}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k}) 
=
\frac{1}{\hbar^2}
\langle n\mathbf{k} |
\big[ r^{\mathrm{a}},[ V^\mathrm{nl}, r^\mathrm{b}]\big]
| m\mathbf{k}\rangle
,
\end{equation} 
and we expand the triple commutator to obtain,
\begin{align*}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k}) 
&=
\frac{1}{\hbar^2\Omega}
\sum_{\mathbf{G},\mathbf{G}'} 
A^*_{n\mathbf{k}}(\mathbf{K}) 
A_{m\mathbf{k}}(\mathbf{K}')
\int
d\mathbf{r} d\mathbf{r}'
 e^{-i\mathbf{K}\cdot\mathbf{r}}
\Big(
r^{\mathrm{a}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^\mathrm{b}
-
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^\mathrm{a}
r'^{\mathrm{b}}
\nonumber\\
&-
r^\mathrm{b}
r^{\mathrm{a}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
+
 r^\mathrm{b}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^{\mathrm{a}}
\Big) 
 e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{align*} 
where 
$V^\mathrm{nl}(\mathbf{r},\mathbf{r}') = \langle\mathbf{r} | V^\mathrm{nl} | \mathbf{r}'\rangle$.
We use the following identity
\begin{align*}
&
\Big(
\frac{\partial^2}{\partial K^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K'^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{a}\partial K^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{b}\partial K'^\mathrm{a}}
\Big)
\int 
d\mathbf{r} d\mathbf{r}' 
 e^{-i\mathbf{K}\cdot\mathbf{r}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
e^{i\mathbf{K}'\cdot\mathbf{r}'}
\nonumber\\
&=
\int d\mathbf{r} d\mathbf{r}'
 e^{-i\mathbf{K}\cdot\mathbf{r}}
\Big( 
r^{\mathrm{a}} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^\mathrm{b}
- 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^\mathrm{a} 
r'^{\mathrm{b}}
- 
r^\mathrm{b} 
r^{\mathrm{a}} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
+
 r^\mathrm{b} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^{\mathrm{a}}
\Big)  
e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{align*}
to write
\begin{equation*}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k})
=
\frac{1}{\hbar^2}
\sum_{\mathbf{G},\mathbf{G}'} 
A^*_{n\mathbf{k}}(\mathbf{K}) 
A_{m\mathbf{k}}(\mathbf{K}')
\Big(
\frac{\partial^2}{\partial K^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K'^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{a}\partial K^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{b}\partial K'^\mathrm{a}}
\Big)
\langle \mathbf{K} |
V^\mathrm{nl} 
| \mathbf{K}'\rangle,
\end{equation*} 
where
\begin{equation*}
\langle\mathbf{K} |
V^\mathrm{nl} 
| \mathbf{K}'\rangle
=
\frac{1}{\Omega}
\int 
d\mathbf{r} d\mathbf{r}' 
 e^{-i\mathbf{K}\cdot\mathbf{r}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{equation*}
The double derivatives with respect to $\mathbf{K}$ and $\mathbf{K}'$ 
can be worked out explicitly and 
$L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k})$
can be finally calculated.\cite{valerie}

>>>>

\subsection{Length Gauge Formalism}
To calculate the optical response 
within the Length Gauge formalism, we take the 
interaction Hamiltonian as
\begin{equation}
H_{I}(t)=-e\mathbf{r}\cdot \mathbf{E}(t).
\label{rde}
\end{equation}
The 
treatment of the position operator
$\mathbf{r}$ for extended Bloch states, $\langle\mathbf{r}|n\mathbf{k}\rangle$ has 
been described in detail.\cite{adamsJCP53,blountSSP62} 
In particular, Ref.~\onlinecite{aversaPRB95} develops
the correct treatment of $\mathbf{r}$ for
the calculation of the nonlinear optical response.
$\mathbf{r}$ is split into the {\it intraband} 
($\mathbf{r}_i$) and {\it interband} ($\mathbf{r}_e$) parts, where 
$\mathbf{r}=\mathbf{r}_i+\mathbf{r}_e$. 
For $\mathbf{r}_e$ one uses
\begin{equation}
\langle n\mathbf{k} | \mathbf{r}_e | m\mathbf{k}'\rangle =
(1-\delta_{nm})\delta(\mathbf{k}-\mathbf{k}')\boldsymbol{\xi}_{nm}(\mathbf{k})
,
\label{rnminn}
\end{equation}
such that $\mathbf{r}_{e,nm}=0$ for $n=m$.
From Eq.~\eqref{mv} with $H_0\to
H^\Sigma_0$, we obtain
\begin{align}
\mathbf{r}_{e,nm}(\mathbf{k}) =
\boldsymbol{\xi}_{nm}(\mathbf{k})\equiv 
\mathbf{r}_{nm}(\mathbf{k}) 
&=
\frac{\mathbf{v}^\Sigma_{nm}(\mathbf{k})}{i\omega^\Sigma_{nm}(\mathbf{k})}
\quad\quad n\notin D_m 
,
\label{pmnrmn}
\end{align}  
where we defined
$\omega^\Sigma_{nm}(\mathbf{k})\equiv\omega^\Sigma_n(\mathbf{k})-\omega^\Sigma_m(\mathbf{k})$, and
$D_m$ are all the possible degenerate $m$-states. 
%We recognize that for each instance in which 
%$n\notin D_m$ or $n\ne m$, $\mathbf{r}_{nm}(\mathbf{k})$ are interband matrix
%elements.
For the intraband part, $\mathbf{r}_i$ only appears in
commutators during the derivation of
the optical response. We use,\cite{aversaPRB95}
\begin{equation}
\langle n\mathbf{k} | [\mathbf{r}_i,{\cal O}] | m\mathbf{k}'\rangle
=i\delta(\mathbf{k}-\mathbf{k}')({\cal O}_{nm})_{;\mathbf{k}},
\label{conmri3n}
\end{equation}  
where
\begin{equation}
({\cal O}_{nm})_{;\mathbf{k}}=
\nabla_\mathbf{k} 
{\cal O}_{n m}(\mathbf{k}) 
- 
i 
{\cal O}_{nm}(\mathbf{k}) 
\left(
\boldsymbol{\xi}_{nn}(\mathbf{k}) 
-
\boldsymbol{\xi}_{mm}(\mathbf{k}) 
\right) 
,
\label{gendevnn}
\end{equation} 
is
the generalized derivative 
of the operator ${\cal O}$. 
The vectors $\boldsymbol{\xi}_{nn}(\mathbf{k})$ are defined in 
Ref.~\onlinecite{aversaPRB95} though they do not need to be 
calculated explicitly in what follows.

@@

In this section we present a 
relevant test case to check the
consistency of our approach. We have selected
a clean Si(001) surface with a $2\times 1$ surface reconstruction.
The slab for such a surface could be chosen to be centrosymmetric 
by creating the front and back
surfaces with the same $2\times 1$
reconstruction. However, we choose to terminate one 
of the surfaces with hydrogen producing an ideal  
terminated bulk Si surface. The H atoms simply saturate the 
dangling bonds of the bulk-like Si atoms at the surface, as seen in 
Fig.~\ref{si2x1}.
We take the $z$ coordinate pointing out of the surface and the $x$
coordinate along the crystallographic [011] direction is parallel to the dimmers. 
\begin{figure}
\centering 
\includegraphics[scale=.8]{images/si2x1-crop}
\caption{(color on line) The slab shows a 
    clean Si(001)$2\times 1$ front surface 
    with an ideal terminated Si bulk back surface. The
    dangling bonds are H (small balls) saturated. 
    This image depicts 12 Si atomic layers
    with one H atomic layer. 
\label{si2x1}} 
\end{figure} 
The idea behind this slab configuration is that the 
cristalline symmetry of the H terminated surface imposes that 
$\chi_{\mathrm{H}}^{xxx}=0$.
 The $2\times 1$ surface has no 
such restrictions, so $\chi_{2\times 1}^{xxx}\ne 0$.
This is due to the fact that along the $y$ direction there is a mirror plane for the
H-saturated surface, whereas 
for the $2\times 1$ surface this mirror is lost as the dimers are
asymmetric along $x$. 
Thus, calculating $\chi^{xxx}$ for the full-slab, or the 
half-slab containing the $2\times 1$ surface\cite{note1}
should yield the same result since the contribution from the H
saturated surface is zero regardless. 
We must check that the following 
relationship is satisfied for this particular slab
\begin{equation*}
\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega) 
=
\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega) 
,
\end{equation*}
where
$\chi_{\mathrm{half-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ from the upper half containing the $2\times 1$ 
surface reconstruction, as seen in Fig.~\ref{si2x1},
and $\chi_{\mathrm{full-slab}}^{xxx}(-2\omega;\omega,\omega)$ is calculated using
${\mathbf{\cal C}}(z)=1$ through the full slab.
We show the results for this comparison in the remainder 
of this section.

@@

From Eqs.~\eqref{lab} and \eqref{vnl} 
\begin{equation*}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k}) 
=
\frac{1}{\hbar^2}
\langle n\mathbf{k} |
\big[ r^{\mathrm{a}},[ V^\mathrm{nl}, r^\mathrm{b}]\big]
| m\mathbf{k}\rangle
,
\end{equation*} 
and we expand the triple commutator to obtain,
\begin{align*}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k}) 
&=
\frac{1}{\hbar^2\Omega}
\sum_{\mathbf{G},\mathbf{G}'} 
A^*_{n\mathbf{k}}(\mathbf{K}) 
A_{m\mathbf{k}}(\mathbf{K}')
\int
d\mathbf{r} d\mathbf{r}'
 e^{-i\mathbf{K}\cdot\mathbf{r}}
\Big(
r^{\mathrm{a}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^\mathrm{b}
-
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^\mathrm{a}
r'^{\mathrm{b}}
\nonumber\\
&-
r^\mathrm{b}
r^{\mathrm{a}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
+
 r^\mathrm{b}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
r'^{\mathrm{a}}
\Big) 
 e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{align*} 
where 
$V^\mathrm{nl}(\mathbf{r},\mathbf{r}') = \langle\mathbf{r} | V^\mathrm{nl} | \mathbf{r}'\rangle$.
We use the following identity
\begin{align*}
&
\Big(
\frac{\partial^2}{\partial K^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K'^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{a}\partial K^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{b}\partial K'^\mathrm{a}}
\Big)
\int 
d\mathbf{r} d\mathbf{r}' 
 e^{-i\mathbf{K}\cdot\mathbf{r}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
e^{i\mathbf{K}'\cdot\mathbf{r}'}
\nonumber\\
&=
\int d\mathbf{r} d\mathbf{r}'
 e^{-i\mathbf{K}\cdot\mathbf{r}}
\Big( 
r^{\mathrm{a}} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^\mathrm{b}
- 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^\mathrm{a} 
r'^{\mathrm{b}}
- 
r^\mathrm{b} 
r^{\mathrm{a}} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}')
+
 r^\mathrm{b} 
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
r'^{\mathrm{a}}
\Big)  
e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{align*}
to write
\begin{equation*}
L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k})
=
\frac{1}{\hbar^2}
\sum_{\mathbf{G},\mathbf{G}'} 
A^*_{n\mathbf{k}}(\mathbf{K}) 
A_{m\mathbf{k}}(\mathbf{K}')
\Big(
\frac{\partial^2}{\partial K^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K'^\mathrm{a}\partial K'^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{a}\partial K^\mathrm{b}}
+
\frac{\partial^2}{\partial K^\mathrm{b}\partial K'^\mathrm{a}}
\Big)
\langle \mathbf{K} |
V^\mathrm{nl} 
| \mathbf{K}'\rangle,
\end{equation*} 
where
\begin{equation*}
\langle\mathbf{K} |
V^\mathrm{nl} 
| \mathbf{K}'\rangle
=
\frac{1}{\Omega}
\int 
d\mathbf{r} d\mathbf{r}' 
 e^{-i\mathbf{K}\cdot\mathbf{r}}
V^\mathrm{nl}(\mathbf{r},\mathbf{r}') 
e^{i\mathbf{K}'\cdot\mathbf{r}'}
,
\end{equation*}
The double derivatives with respect to $\mathbf{K}$ and $\mathbf{K}'$ 
can be worked out explicitly and 
$L^{\mathrm{a}\mathrm{b}}_{nm}(\mathbf{k})$
can be finally calculated.\cite{valerie}
